---
title: "2026 Student Notebook - part 2"
author: "Student 1, Student 2, Student 3"
output: pdf_document
date: "2026-05-12"
---

## Student Self-Assessment

Data Science - University of Deusto

Before submission, provide an estimation of:

-   **Total time spent:** [Insert time]
-   **Main difficulties:** [Insert text]
-   **Level of satisfaction:** [Insert text]
-   **Improvement suggestions:** [Insert text]

------------------------------------------------------------------------

## Context of the project and instructions

This is the notebook you will use for the assignment of the subject Data Science.

You will find here exercises regarding all the topics covered within the semester. The exercises are aligned with theory classes and examples of code you will see during sessions, so it is recommendable to update it weekly in order to not accumulate all of them at the end of the semester.

Please **read carefully the statements of the different exercises**. If you have any question, please use the forum (but do not share code).

## Submission of the Notebook

To submit the notebook: - Print it in a **PDF file** with all the sections expanded and all the cells executed, in a way all the code can be seen, and all the results are shown. - Save the notebook as **.Rmd file**

Submit both files (NOT in a zip folder) into ALUD.

## Evaluation of the Notebook

The evaluation of the Notebooks counts 50% of the final qualification of the subject. Evaluation is divided into two notebooks (25% each).

The weight in the final mark of all the exercises is equal.

In some exercises, there are extra instructions indicated with the word **(Challenge)**. These are not requirements, but actions needed to surpass the global qualification of 8.5 points.

### Grading Criteria:

-   **[8.5 to 10] points**: Excellent solution; meets all requirements and **(Challenge)**'s.
-   **[5 to 8.5] points**: Meets requirements but has room for improvement (presentation, justification, code efficiency).
-   **[1 to 5] points**: Does not meet several requirements.
-   **[0] points**: Solution not provided or plagiarised.

### Assessment Breakdown:

-   **Functionality (50%)**: The code must meet requirements and run without errors.
-   **Clarity and Presentation (50%)**: Includes structure, readability, explanations, and proper display of results.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(glue)
library(knitr)
library(dbscan)
library(FactoMineR)
library(factoextra)
library(tidytext) # Package necessary for the stop_words
```

## Exercise 01: Text mining

*Using dataset: "Recipe Reviews"*

1.  Clean the `text` column of the reviews: convert to lowercase, remove punctuation, and remove numbers. Tokenize the text and remove standard English "stop words".
2.  Generate a comparative frequency table: obtain the top 15 most frequent words used in **positive reviews** (`stars` = 5) versus those used in **negative reviews** (`stars` \<= 2).
3.  Create a "Word Cloud" for the reviews of the recipe that has the highest number of `thumbs_up` in the entire dataset.
4.  **(Challenge):** Perform a basic sentiment analysis by assigning a sentiment score to each comment using a dictionary (e.g., `bing` or `afinn`). Show if there is a correlation between the sentiment score you calculated and the `stars` given by the user.

### Task 1:

```{r, message=FALSE, warning=FALSE}
rm(list = ls()) #This command is included to avoid confusion with previously declared variables
recipeData <- read.csv("Recipe Reviews and User Feedback Dataset.csv") # Load data

data("stop_words") # Load the stop_words dataset

# Clean the reviews
reviews <- recipeData %>%
  mutate(text_clean = tolower(text)) %>%  # Change all the words to lowercase
  mutate(text_clean = str_replace_all(text_clean, "[[:punct:]]", "")) %>% # Eliminate all the punctuation marks from the text
  mutate(text_clean = str_replace_all(text_clean, "[[:digit:]]", "")) %>% # Eliminate all the numbers from the text
  unnest_tokens(word, text_clean) %>% # Tokenize the text to have the words only
  anti_join(stop_words, by = "word") # Anti join to have only the words that are not on the stop words list
knitr::kable(head(reviews$word, 15))
```

#### Results Analysis:

The text preparation and cleaning phase transformed unstructured raw review strings into standardized individual tokens optimized for statistical analysis.

Inspecting the first 15 generated tokens (including tweaked, removed, onions, onion, seasoning, oregano, paprika, cayenne), it can be concluded the following methodological and analytical insights:

* **Filtering Efficiency:** The removal of punctuation marks, numbers, and standard English grammatical stop words (such as the, and, to) worked as intended. This ensures that the final dataset is predominantly populated by culinary nouns (onions, seasoning, oregano, paprika, cayenne) and action-oriented verbs (tweaked, removed, mix), which is the expected vocabulary for the dataset.

* **Recipe Body Dynamics:** The output reflects the behavioral nature of the platform's users. Terms like "tweaked" and "removed" indicate that home cooks frequently modify recipes to fit personal preferences. Furthermore, the high density of spices appearing in the initial preview suggests that the dataset contains heavily seasoned dishes or discussions regarding flavor adjustments.

* **Text Mining Limitations:** The output also shows a few common quirks of basic text cleaning. For instance, because the code doesn't merge variations of the same word, singular and plural forms coexist ("onion" and "onions"), which splits their count. Additionally, a random "x" appears (likely left behind from expressions like "2x" after the numbers were deleted), and general words like "bit" or "called" slip through because standard dictionaries don't know they aren't useful for recipe analysis.

#### Task 2

```{r, message=FALSE, warning=FALSE}
# Using the variable 'reviews' that is clean and tokenized

topPos <- reviews %>%
  filter(stars == 5) %>% # Select only 5 star reviews
  group_by(word) %>% # Group by word to have the amount of words
  summarise(occurrences = n()) %>% # Count the amount of occurrences for each word
  arrange(desc(occurrences)) %>% # Arrange them by the decreasing order of the total number of words
  head(.,15)

topNeg <- reviews %>% 
  filter(stars <= 2) %>% # Select only 2 star or lower reviews
  group_by(word) %>% # Group by word to have the amount of words
  summarise(occurrences = n()) %>% # Count the amount of occurrences for each word
  arrange(desc(occurrences)) %>% # Arrange them by the decreasing order of the total number of words
  head(.,15)

# Display frequency results for comparison
frequencyTable <- cbind(topPos, topNeg)
knitr::kable(frequencyTable)
```

#### Results Analysis:

The comparative table displays the top 15 most frequent words used in positive reviews (5 stars) versus negative reviews (<= 2 stars). Comparing these two lists reveals some very interesting insights about how users leave feedback:

* **Massive Volume Difference:** The first thing that stands out is the sheer scale of occurrences. The top words in positive reviews appear thousands of times (e.g., "recipe" has 7,757 hits), while the top words in negative reviews barely touch a few hundred (e.g., "recipe" has 997 hits). This tells us that the dataset contains a much larger number of satisfied users leaving 5-star reviews than unhappy ones.

* **Shared Vocabulary (The Context):** Interestingly, both positive and negative reviews share a lot of the same core words, such as "recipe", "cake", "time", "added", "family", and ingredients like "cheese" and "bread". This makes sense because no matter if a cook loves or hates a dish, they are still talking about the same activity: baking a cake or making dinner for their family.

* **The Subtle Differences in Tone:** Even though the lists look similar, the subtle changes tell the real story. In the positive column, enthusiastic words like "delicious" (2,724 hits) and "loved/love" score very high. In contrast, the negative column includes words like "sugar", "flour", and "taste". This suggests that when reviews are bad, users tend to focus less on compliments and more on specific ingredients or complaining about how the final dish actually tasted.

* **Positive Words in Negative Reviews:** It might seem strange that "love" and "delicious" appear in the negative column. This could happen because of negations (like "I did not love it") or mixed feedback (like "It smelled delicious, but tasted awful"). Since the code only counts single words, it misses the negative context.

In short, while the general topics of conversation remain the same across all ratings, positive reviews are much more vocal and loaded with emotional praise, whereas negative reviews lean toward breaking down what went wrong with the food.

#### Task 3

```{r, message=FALSE, warning=FALSE}
# Identify the recipes with the highest engagement based on total thumbs up
topRecipesId <- recipeData %>%
  group_by(recipe_code) %>%
  summarise(total_thumbs = sum(thumbs_up)) %>%
  arrange(desc(total_thumbs))

# Filter words from reviews corresponding only to the top recipe for the word cloud
wordCloud <- reviews %>%
  filter(recipe_code == topRecipesId$recipe_code[1]) %>%
  count(word, sort = TRUE)

set.seed(123) # Seed to have always the same answers
library(wordcloud)
wordcloud(
  words = wordCloud$word, 
  freq = wordCloud$n, 
  max.words = 100,               # Shows the 100 most common words
  random.order = FALSE,          # Leaves the biggest words in the centre
  colors = brewer.pal(8, "Dark2") # Una paleta de colores limpia y legible
)
```

#### Results Analysis:

The word cloud visualizes the 100 most frequent words from the reviews of the single recipe with the highest user engagement (measured by the total sum of thumbs_up). This visual representation allows us to reverse-engineer the identity and reception of the dish:

* **Identifying the Recipe:** Even without looking at the official recipe name, the core ingredients and keywords tell us exactly what this dish is. The prominent size of words like "breakfast", "bacon", "sausage", "cheese", "eggs", and "casserole" reveals that the top recipe in the dataset is a savory breakfast or brunch bake. Other supporting ingredients like "potatoes", "ham", and "swiss" (cheese) help map out the exact profile of the meal.

* **Context of Consumption:** The cloud highlights when and how people enjoy this dish. Words like "morning", "brunch", "christmas", "overnight", and "guests" pop up frequently. This indicates that users often prepare this specific casserole for special family gatherings, holiday mornings, or when hosting company, making it a reliable crowd-pleaser.

* **Overwhelmingly Positive Sentiment:** The overall feedback embedded in the cloud is highly enthusiastic. Terms such as "delicious", "loved", "easy", "perfect", and "wonderful" are centrally placed and quite large. This confirms that its high engagement status is driven by genuine popularity and success among home cooks, rather than controversial or bad reviews.

### Challenge:

```{r, message=FALSE, warning=FALSE}
# get_sentiments("bing") is a dictionary with sentiment words with a category (negative/positive)
sentimentsBing <- get_sentiments("bing")

# Calculate the net sentiment score for each unique comment
score_Comments <- reviews %>%
  inner_join(sentimentsBing, by = "word") %>%
  mutate(value = ifelse(sentiment=="positive", 1, -1)) %>%
  group_by(comment_id, stars) %>%
  summarise(sentiment_score = sum(value), .groups = "drop")

# Compute Pearson correlation coefficient between actual user stars and calculated sentiment score
correlation <- cor(score_Comments$sentiment_score, score_Comments$stars, method = "pearson")

cat("Pearson's correlation coeficient:", correlation, "\n")

# Visualize the relationship using boxplots
ggplot(score_Comments, aes(x = as.factor(stars), y = sentiment_score, fill = as.factor(stars))) +
  geom_boxplot(show.legend = FALSE) +
  labs(
    title = "Relation Between Stars and the Sentiment Score",
    x = "Stars",
    y = "Sentiment Score"
  ) +
  theme_minimal()
```

#### Results Analysis:

This section shows the relationship between the computed sentiment_score (using the Bing dictionary) and the explicit stars given by the users. The relationship has a Pearson correlation coefficient of 0.1849, which indicates a weak positive linear correlation.

While a higher star rating does move the sentiment score in a positive direction, the correlation is much lower than one might intuitively expect. Analyzing the boxplot reveals the core reasons behind this weak statistical relationship:

* **General Upward Trend:** Looking at the medians (the thick black lines inside the boxes) from 1 star to 5 stars, there is a clear, gradual upward steps. One-star reviews have a negative median score (around -1), while five-star reviews settle into positive territory. This proves that the dictionary is capturing the basic emotional signal correctly.

* **The Zero-Star Anomaly:** Interestingly, reviews marked with "0" stars show a positive median score close to 1. This suggests that these entries might represent unrated reviews rather than genuinely terrible feedback, skewing the overall linear calculation.

* **High Variance and Outliers:** The most striking feature of the plot is the massive spread of data and the heavy presence of outliers (the black dots), especially in the 4 and 5-star categories. For instance, in the 5-star column, there are multiple reviews with computed sentiment scores dropping as low as -5 to -9.

* **Methodological Limitations:** This heavy variance highlights why simple, word-by-word (unigram) sentiment analysis struggles in real-world contexts. A user might write a 5-star review saying: "There is no way this recipe could be bad, it saved me from a total kitchen disaster." A basic dictionary counts the negative words "no", "bad", and "disaster", giving a deeply negative score for a highly enthusiastic review. It completely misses context, intensifiers, and double negatives.

In conclusion, while the sentiment model successfully establishes a positive baseline trend that aligns with user ratings, a Pearson coefficient of 0.1849 shows that counting isolated words provides an incomplete picture. Human emotion, sarcasm, and complex sentences introduce a high level of noise that simple dictionary matching cannot fully filter out.

## Exercise 02: Table Transformations and Time Series

*Using dataset: "retail.csv"*

1.  Ensure `InvoiceDate` has the correct date/time format. Extract the hour of the day (`Hour`) and the day of the week (`Weekday`).
2.  Pivot the table to obtain a Heatmap (or cross-tabulation) where rows are days of the week, columns are hours of the day (0 to 23), and the cell value is the **total revenue** (`Quantity * UnitPrice`). Interpret: What is the peak sales time?
3.  Transform the dataset to obtain a summary table per **Customer** (`CustomerID`) containing:

-   **Recency:** Days elapsed since their last purchase to the maximum date in the dataset.
-   **Frequency:** Number of unique invoices (`InvoiceNo`).
-   **Monetary:** Total accumulated spending.

```{r, message=FALSE, warning=FALSE}
rm(list = ls())
options(scipen = 999)

##########
# TASK 1 #
##########

data <- read.csv("retail.csv")
data$InvoiceDate <- ymd_hms(data$InvoiceDate)

data <- data %>%
  mutate(
    Hour = hour(InvoiceDate),
    WeekDay = weekdays(InvoiceDate)
  )

```

```{r}
##########
# TASK 2 #
##########

dataRevenue <- data %>%
  group_by(WeekDay, Hour) %>%
  summarise(
    Revenue = sum(Quantity * UnitPrice)
  )

dataWider <- dataRevenue %>%
  pivot_wider(
    names_from = Hour,
    values_from = Revenue
  )

ggplot(dataRevenue, aes(x = Hour, y = WeekDay, fill = Revenue)) +
  geom_tile()

# We can see peak revenue Thursdays at 10 am
```

```{r}
##########
# TASK 3 #
##########

dataRFM <- data %>%
  group_by(CustomerID) %>%
  summarise(
    Frequency = n_distinct(InvoiceNo),
    Monetary = sum(Quantity * UnitPrice),
    Recency = as.numeric(difftime(max(data$InvoiceDate), max(InvoiceDate), units = "days"))
  )
```

#### Results Analysis:

The heatmap reveals that peak revenue occurs on Thursday around 9–10h, standing out clearly from all other time slots. More broadly, Tuesday through Thursday show consistently high revenue between 9h and 15h. Saturday is entirely absent from the chart, and Sunday shows only minimal activity, confirming the business operates almost exclusively on weekdays. Revenue drops sharply before 8h and after 18h across all days, indicating no relevant demand outside standard working hours.

## Exercise 03: Clustering

*Using the RFM table generated in Block 2 (Retail)*

1.  Normalize the `Recency`, `Frequency`, and `Monetary` variables calculated in the previous exercise.
2.  Determine the optimal number of clusters using the Elbow Method or Silhouette Coefficient.
3.  Apply the **K-Means** algorithm with the chosen number of clusters. Add the cluster assignment to the original customer table. 
4.**(Challenge):** Profile the obtained clusters: Calculate the centroid (mean of R, F, M) for each group and assign a **commercial label** to each one (e.g., "VIP Customers", "At Risk", "New", etc.). Justify your labeling based on the data.

```{r, message=FALSE, warning=FALSE}
##########
# TASK 1 #
##########

dataRFMNorm <- scale(dataRFM[,-1])

```

```{r}
##########
# TASK 2 #
##########

fviz_nbclust(dataRFMNorm, kmeans, method = "wss", k.max = 20) + 
  ggtitle("Optimal number of clusters")

```

```{r}
##########
# TASK 3 #
##########

k <- 4
km.out_1 <- kmeans(dataRFMNorm, 
                   k, 
                   nstart = 10 # how many initial conditions are created
)

plot(dataRFMNorm, col = (km.out_1$cluster + 1),
     main = paste("K-Means Clustering Results (Original) with K = ", k, sep = " "),
     xlab = "", ylab = "", pch = 20, cex = 2)

dataRFM <- dataRFM %>%
  mutate(
    Cluster = km.out_1$cluster
  )

```

```{r}
##########
# TASK 4 #
##########

dataRFM <- dataRFM %>%
  group_by(Cluster) %>%
  summarise(
    Recency = mean(Recency),
    Frequency = mean(Frequency),
    Monetary = mean(Monetary)
  )

dataRFM <- dataRFM %>%
  mutate(Label = case_when(
    Cluster == 1 ~ "VIP Customers",
    Cluster == 2 ~ "At Risk",
    Cluster == 3 ~ "Loyal Customers",
    Cluster == 4 ~ "Lost Customers"
  ))

# Cluster 1 (VIP) shows the lowest recency (4 days) with by far the highest frequency (130 invoices) and monetary value (€132k), indicating highly active and valuable customers. Cluster 4 (Lost) shows the opposite: last purchase was 248 days ago with minimal frequency and spending, suggesting churned customers. Clusters 2 and 3 fall in between, differentiated by recency and engagement level.
```

## Exercise 04: Association Rules

*Using dataset: "retail.csv"*

1.  **Transactional Preparation:** Using `retail.csv`, group the data to obtain a transaction structure ("Basket") where each row represents an `InvoiceNo` and contains the list of `Description` (products) purchased. Remove transactions containing only a single item.
2.  **Rule Generation:** Apply the **Apriori** algorithm. Adjust the *Support* and *Confidence* parameters to obtain a set of between 20 and 100 rules.
3.  **Filtering:** List the 5 rules with the highest **Lift**. Interpret in your own words what the top rule implies (e.g., "If a customer buys X, they are Y times more likely to buy Z...").
4.  **(Challenge):** Filter the rules to show only those where the **consequent (RHS)** is a specific high-rotation product (choose a frequent one, e.g., "WHITE HANGING HEART..."). Visualize these relationships using a graph (using `arulesViz` or similar).

```{r, message=FALSE, warning=FALSE}
rm(list = ls()) 

##############################
#1 Transactional Preparation #
##############################
library(arules)

data <- read.csv("retail.csv")
#Change the structure
data_single <- data %>%
  select(InvoiceNo, Description) %>%
  drop_na(Description) %>%
  distinct(InvoiceNo, Description)
#Change to transactions format
transactions <- as(split(data_single$Description, data_single$InvoiceNo), "transactions")
#Apply condition
transactions <- transactions[size(transactions) > 1]
```

The dataset was transformed into transactions, where each InvoiceNo represents one basket containing the purchased products. Transactions with only one product were removed because association rules require relationships between multiple items.

```{r}
####################
#2 Rule Generation #
####################

#Apply Apriori algorithm
rules <- apriori(transactions, parameter = list(supp = 0.002, conf = 0.3))
length(rules)
#Number of rules: 11583161

#If there are too many rules:
#Increase support/confidence -> fewer rules
#Decrease support/confidence -> more rules

#If there are too few rules:
#Decrease support/confidence -> more rules
#Increase support/confidence -> fewer rules

rules <- apriori(transactions, parameter = list(supp = 0.006, conf = 0.4))
length(rules)
#Number of rules: 9742

rules <- apriori(transactions, parameter = list(supp = 0.01, conf = 0.5))
length(rules)
#Number of rules: 808

rules <- apriori(transactions, parameter = list(supp = 0.015, conf = 0.6))
length(rules)
#Number of rules: 97

rules <- apriori(transactions, parameter = list(supp = 0.016, conf = 0.65))
length(rules)
#Number of rules: 56 -> is between 20 and 100, "in the middle"

#View rules
inspect(rules[1:10])
```

The Apriori algorithm was applied to discover associations between products. Different support and confidence values were tested to control the number of generated rules. Finally, support = 0.016 and confidence = 0.65 generated 56 rules, which satisfies the requested range between 20 and 100 rules.

```{r}
##############
#3 Filtering #
##############

#Variable for the Top5
Top5_rules <- sort(rules, by = "lift", decreasing = TRUE)[1:5]
#Show
inspect(Top5_rules)

#LHS -> First product
#RHS -> Second product
#confidence -> buy LHS and then RHS
#lift -> buy RHS and then LHS
```

If a customer buys "REGENCY TEA PLATE GREEN", they are about 36.9 times more likely to also buy "REGENCY TEA PLATE ROSES" compared to a random customer. In addition, around 83.6% of customers who buy the green plate also buy the roses plate.

```{r}
##############
#4 Challenge #
##############

#Load library
library(arulesViz)

#Check frequent products
sort(itemFrequency(transactions), decreasing = TRUE)[1:10]

#Filter rules where RHS is WHITE HANGING HEART T-LIGHT HOLDER
rules_filtered <- subset(
  rules,
  rhs %in% "WHITE HANGING HEART T-LIGHT HOLDER"
)

#Check filtered rule
inspect(rules_filtered)

#Plot graph
plot(rules_filtered, method = "graph")
```

The rules were filtered using a specific RHS product. The graph visualization shows the association between the selected product and the products that are commonly purchased together with it.